Visual Analytics of Large Bipartite Networks assisted by Multilevel Strategies\thanks{Supported by FAPESP, project 2017/05838-3}

### 2019-05-22 paper.tex
feito ate os conceitos fundamentais

\begin{document}
%
\title{Visual Analytics of Large Bipartite Networks assisted by Multilevel Strategies\thanks{Supported by FAPESP, project 2017/05838-3}}
%
\titlerunning{Multilevel Visualization of Bipartite Networks}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Renato Fabbri\orcidID{0000-0002-9699-629X} \and
Alan Valejo \and
Maria Cristina Ferreira de Oliveira\orcidID{0000-0002-4729-5104}}
%
\authorrunning{R. Fabbri et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{University of São Paulo, São Carlos SP, BR\\
\email{renato.fabbri@gmail.com},
\email{alanvalejo@gmail.com},
\email{cristina@icmc.usp.br}\\
\url{http://conteudo.icmc.usp.br/pessoas/cristina/}}
%
\maketitle
%
\begin{abstract}
% The abstract should briefly summarize the contents of the paper in
% 150--250 words.
  It is a well established fact that bipartite, or two-layer, networks are pervasive
  in model real-world phenomena and that they play fundamental roles in
  graph theory.
  Multilevel strategies have been developed for optimization tasks,
  and for the visualization of simple (``unipartite'') networks,
  but their employment for visualizing bipartite networks were not found by the authors.
  In this work, we present advances in the use of multilevel strategies for the
  visualization of bipartite networks,
  allowing interactive and intuitive navigation of such structures and visual mappings of large datasets.
  More specifically, we developed a visual analytics web interface in which
  a parametrizable simpification of
  bipartite networks are obtained through the application of coarsening algorithms.
  The resulting networks are then presented to the user,
  providing a genuine route for the ``overview first - focus on demand''
  process on the analysis of the undelying data, in which the analyst
  selects supervertices or whole network sectors for more detailed observation,
  i.e. performs requests for the interface to display 
  specific structures in less simplified settings.
  Moreover, the application is useful for the development multilevel strategies
  e.g. by the specification of vertices to guide the coarsening processes and
  the examination of the resulting multilevel hierarchy.

\keywords{Network visualization \and Multilevel strategies \and Visual analytics \and Big data \and Complex networks \and Data visualization.}
\end{abstract}
%
\section{Introduction}
The visualization of large-scale networks poses challenges both in terms of computational costs
and of effective presentation of the information for the user~\cite{tang,staudt}.
These issues may be aggravated in the case of bipartite networks,
due to their sparsity and topological complexities~\cite{alan1}.
Bipartite networks are comprised of two partitions of nodes,
called ``layers'',
and links are not incident between nodes in the same partition.
Such network type arises very often and naturally from the representation
of relations among two kinds objects,
e.g. documents and terms or authors~\cite{doc,sci,movie}, or patient and gene~\cite{gene}.
Furthermore, real-world networks are often bipartite, and most unipartite networks
are projections of bipartite networks or may be considered as exibiting bipartite properties~\cite{guillaume0,guillaume}.
In order to assist the visualization and navigation of large networks, one possibility is the use of
multilevel strategies, which consist on the employment of incremental coarsening of the original
network to obtain a sequence of simplified representations.
Multilevel strategies are most traditionally used for executing complex algorithms
on large-scale networks by the application of the algorithm on a smaller
version of the network~\cite{alan2,ml2}.
Their employment of multilevel strategies for the visualization of simple (i.e. ``unipartite'')
networks have been reported, but their exploitation for visualizing bipartite networks
was not found by the authors.
Accordingly, we present a system for the visualization of bipartite networks using
multilevel strategies developed for bipartite networks.
The system consists on presented a simplified version of the network for the user, which then
requests for supervertices (or collections of them) to be uncoarsened and presented in more detail.

This paper is organized as follows: in Section~\ref{rel} the related work is examined, while in~\ref{nom} are selected remarks about the vocabulary.
Fundamental concepts are introduced in Section~\ref{bac}, and the method is delineated in Section~\ref{des}.
The software implementation is then described in Section~\ref{sof}.
Results and discussion are in~\ref{res}.
Finally, Section~\ref{con} holds concluding and further work statements.

\subsection{Related work}\label{rel}
%%%
% artigos que devemos citar pelo histórico científico
% da estratégia ML, da visualização ML
% e os que devemos citar pelo ML em redes bipartidas
% e desenvolvidos no ICMC
Multilevel strategies have been employed to visualize unipartite networks~\cite{u1,u2,u3,u4,u5,u6,u7}.
Also, the aggregation of clusters have been reported, and comprises an approach that resembles
the coarsening procedure in creating simplified representations of the original network~\cite{a1,a2,a3,a4,a5,a6}.
Even so, the authors are not aware of previous reports on the use of multilevel strategies
for the visualization and navigation of bipartite networks.

\subsection{Nomenclature and conceptual remarks}\label{nom}
%%%
% level vs layer
% simple and heterogeneous graphs
% multiplex, ?
% multilevel strategies for optimization
% supernode metanode,

\section{Backgroud}\label{bac}
\section{Method description}\label{des}
%%%
% mlpb and other methods developed
% arguments for the preferred method (mlpb?)
\begin{figure}[!h]\centering
% \includegraphics[width=.7\textwidth]{}
  \caption{.
  }\label{fig:glob}
\end{figure}

>>>>

\begin{document}
%
\title{Visual Analytics of Large Bipartite Networks assisted by Multilevel Strategies\thanks{Supported by FAPESP, project 2017/05838-3}}
%
\titlerunning{Multilevel Visualization of Bipartite Networks}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Renato Fabbri\orcidID{0000-0002-9699-629X} \and
Alan Valejo \and
Maria Cristina Ferreira de Oliveira\orcidID{0000-0002-4729-5104}}
%
\authorrunning{R. Fabbri et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{University of São Paulo, São Carlos SP, BR\\
\email{renato.fabbri@gmail.com},
\email{alanvalejo@gmail.com},
\email{cristina@icmc.usp.br}\\
\url{http://conteudo.icmc.usp.br/pessoas/cristina/}}
%
\maketitle
%
\begin{abstract}
% The abstract should briefly summarize the contents of the paper in
% 150--250 words.
  It is a well established fact that bipartite, or two-layer, networks are pervasive
  in model real-world phenomena and that they play fundamental roles in
  graph theory.
  Multilevel strategies have been developed for optimization tasks,
  and for the visualization of simple (``unipartite'') networks,
  but their employment for visualizing bipartite networks were not found by the authors.
  In this work, we present advances in the use of multilevel strategies for the
  visualization of bipartite networks,
  allowing interactive and intuitive navigation of such structures and visual mappings of large datasets.
  More specifically, we developed a visual analytics web interface in which
  a parametrizable simplification of
  bipartite networks are obtained through the application of coarsening algorithms.
  The resulting networks are then presented to the user,
  providing a genuine route for the ``overview first - focus on demand''
  process on the analysis of the underlying data, in which the analyst
  selects supervertices or whole network sectors for more detailed observation,
  i.e. performs requests for the interface to display 
  specific structures in less simplified settings.
  Moreover, the application is useful for the development multilevel strategies
  e.g. by the specification of vertices to guide the coarsening processes and
  the examination of the resulting multilevel hierarchy.

\keywords{Network visualization \and Multilevel strategies \and Visual analytics \and Big data \and Complex networks \and Data visualization.}
\end{abstract}
%
\section{Introduction}
The visualization of large-scale networks poses challenges both in terms of computational costs
and of effective presentation of the information for the user~\cite{tang,staudt}.
These issues may be aggravated in the case of bipartite networks,
due to their sparsity and topological complexities~\cite{alan1}.
Bipartite networks are comprised of two partitions of nodes,
called ``layers'',
and links are not incident between nodes in the same partition.
Such network type arises very often and naturally from the representation
of relations among two kinds objects,
e.g. documents and terms or authors~\cite{doc,sci,movie}, or patient and gene~\cite{gene}.
Furthermore, real-world networks are often bipartite, and most unipartite networks
are projections of bipartite networks or may be considered as exhibiting bipartite properties~\cite{guillaume0,guillaume}.
In order to assist the visualization and navigation of large networks, one possibility is the use of
multilevel strategies, which consist on the employment of incremental coarsening of the original
network to obtain a sequence of simplified representations.
Multilevel strategies are most traditionally used for executing complex algorithms
on large-scale networks by the application of the algorithm on a smaller
version of the network~\cite{alan2,ml2}.
Their employment of multilevel strategies for the visualization of simple (i.e. ``unipartite'')
networks have been reported, but their exploitation for visualizing bipartite networks
was not found by the authors.
Accordingly, we present a system for the visualization of bipartite networks using
multilevel strategies developed for bipartite networks.
The system consists on presented a simplified version of the network for the user, which then
requests for supervertices (or collections of them) to be uncoarsened and presented in more detail.

This paper is organized as follows: in Section~\ref{rel} the related work is examined, while in~\ref{nom} are selected remarks about the vocabulary.
The method is delineated in Section~\ref{des} and
the software implementation is then described in Section~\ref{sof}.
Results and discussion are in~\ref{res}.
Finally, Section~\ref{con} holds concluding and further work statements.

\subsection{Related work}\label{rel}
%%%
% artigos que devemos citar pelo histórico científico
% da estratégia ML, da visualização ML
% e os que devemos citar pelo ML em redes bipartidas
% e desenvolvidos no ICMC
Multilevel strategies have been employed to visualize unipartite networks~\cite{u1,u2,u3,u4,u5,u6,u7}.
Also, the aggregation of clusters have been reported, and comprises an approach that resembles
the coarsening procedure in creating simplified representations of the original network~\cite{a1,a2,a3,a4,a5,a6}.
Even so, the authors are not aware of previous reports on the use of multilevel strategies
for the visualization and navigation of bipartite networks.
Furthermore, only~\cite{a1,a2,a3,a4,a6} reports on a system for visual analytics, i.e. the other works do not
address data analysis by interactive visual interfaces,
including all multilevel strategies for visualization.

\subsection{Nomenclature remarks}\label{nom}
%%%
% level vs layer
% simple and heterogeneous graphs
% multiplex, ?
% multilevel strategies for optimization
% supernode metanode, 
The main vocabulary issue that arises in the context of this article is between
level and layer. Layers are the sets of a bipartite network, while levels are the sequence
of coarsened networks.
Also, simple or homogeneous networks are opposed to heterogeneous networks, in which there
are more then one type of node.
Bipartite networks may be regarded as an elementary case of heterogeneous networks,
but it has one additional restriction: only nodes of different types are connected.
Another usual concept in this context is that of a multilayer network, in which there are layers,
i.e. partitions of nodes, in addition to nodes and links.
Bipartite networks, in this case, are multilayer networks with only two layers and no intra-layer links, i.e. all the links are inter-layers.
Most importantly, there are many synonyms which are found in this context, e.g. supernodes are also called metanodes or supervertices.
A thorough exposition of the vocabulary are beyond the scope of this article, but
some attention for the issue is helpful to assist searches and newcomers.

\section{Method description}\label{des}
%%%
% mlpb and other methods developed
% arguments for the preferred method (mlpb?)
\subsection{Fundamental concepts}\label{bac}
A bipartite network $G=(V,E)$ consists in a set $V$ of vertices which is partitioned in two subsets
with no links between nodes in the same set, i.e. $\exists\; V_1, V_2:\; V_1\cup V_2 = V,\;V_1\cap V_2 = \emptyset$, and $E\subseteq V_1 \times V_2$.
The variable names are borrowed from more traditional Graph theory, with Vertices (nodes) and Edges (links).

A multilevel strategy consists in obtaining a hierarchy of coarsened networks $G_l$, $l$ integer and $l \in [0,L-1]$ where $G_0$ is the original network.
The coarsening procedures requires two algorithms, the \emph{matching}, that defines the nodes to be collapsed, and the \emph{contraction}, which builds the reduced representation
from the matched nodes.
There are several coarsening algorithms reported in the scientific literature~\cite{}, a few of them
developed for bipartite networks~\cite{}.
A discussion of these algorithms is found on the bibliography~\cite{} and is beyond the scope of this article.
Most importantly, we are interested in the MLPB algorithm, not only because it is suited for bipartite networks, but because it has yielded simplified networks which carries essential topological features of the original networks, and that result is perceived is the visualization as shown in Sections~\ref{des},~\ref{sof} and~\ref{res}.

Visual analytics is the scientific field dedicated to analytical reasoning assisted by interactive visual interfaces~\cite{}.
Therefore, the area is specially concerned with coupling interactive visual representations with sense and decision making.
Of special relevance for the present work, the techniques are most often employed to amplify human capabilities in specific ways, which includes reducing the search space, enhancing the recognition of patterns and the infererence of relationships, and providing a manipulable medium for the exploration of the information of interest~\cite{}.

\subsection{The navigation pathway}

\begin{figure}[!h]\centering
% \includegraphics[width=.7\textwidth]{}
  \caption{.
  }\label{fig:glob}
\end{figure}
